Notes on \\ Projected Knowledge Compilation \\ \today

### 2023-08-03 doc/pkc-notes.tex
Polishing

\newcommand{\algo}[1]{\textsf{#1}}
\newcommand{\kc}{\algo{KC}}
\newcommand{\pkc}{\algo{PKC}}
\newcommand{\traverse}{\algo{Traverse}}
\newcommand{\posop}{\algo{Pos}}
\newcommand{\negop}{\algo{Neg}}
\newcommand{\bcp}{\algo{BCP}}
\newcommand{\uprop}{\algo{Uprop}}
\newcommand{\simpl}{\also{Simpl}}

@@

  Many tools for reasoning about Boolean formulas operate on formulas in \emph{conjunctive normal form}, defined as follows.
  A formula is in  \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  We use symbol $\phi$ to denote CNF formulas.

@@

@@

  For Boolean formula $\theta$ and ring weight function $w$, its {\em data ring evaluation} computes
  \begin{eqnarray}
    \drep(\theta, w) & = & \sum_{\alpha \in \dmodelset(\theta)} \;\; \prod_{\lit \in \alpha} w(\lit) \label{eqn:rep}
  \end{eqnarray}
where $\sum$ and $\prod$ represent the ring sum $\radd$ and ring product $\rmul$ operations, respectively.

For a partitioned-operation formula $\psi$ such that $\dependencyset(\psi) \subseteq \dvarset$, its data ring evaluation can be computed based on the recursive structure of the formula:
\begin{itemize}
\item Constants: $\drep(\tautology) = \mulident$ and $\drep(\nil) = \addident$
\item Literals: $\drep(x) = w(x)$ and $\drep(\obar{x}) = \mulident - x$
\item Product operations: 
$\drep(P(\psi_1, \psi_2, \ldots, \psi_k)) = \prod_{1 \leq i \leq k} \drep(\psi_i)$
\item Sum Operations
$\drep(S(\psi_1, \psi_2, \ldots, \psi_k)) = \sum_{1 \leq i \leq k} \drep(\psi_i)$
\item Negations: $\drep(\neg \psi) = \mulident -\psi$
\end{itemize}

@@

We can build on existing knowledge compilers to implement a projected knowledge compiler.

@@

Let $U(\phi)$ denote the set of \emph{unit literals} in $\phi$, i.e., the set of all literals $\lit$ such that clause $\{ \lit \} \in \phi$.  In the following $L$ denotes a set of literals.

We define the following operations for transforming clausal formula $\phi$ into a new clausal formula:
\begin{description}
\item[$\posop(\phi, L)$:] Remove clauses satisfied by the literals in $L$: $\posop(\phi, L) = \{ C \in \phi \mid C \cap L = \emptyset\}$.

\item[$\negop(\phi, L)$:] Remove the negations of the literals in $L$ from all clauses.  Let $\negop(C, L) = \{ \lit \in C \mid  \obar{\lit} \not \in L\}$.
  Then $\negop(\phi, L) = \{ \negop(C, L) \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.

\item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals in $L$, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
$\simplify{\phi}{L} = \posop(\negop(\phi, L), L)$.

@@

\item[$\ureduce{\phi}{x}$:] Express the universal quantification of $\phi$ with respect to variable $x$ in clausal form:
$\ureduce{\phi}{x} = \negop(\phi, \{x, \obar{x}\})$.

@@

When constructing a POF via a bottom-up process, we can perform simplifications as the formula is constructed, as follows:
\begin{itemize}
\item
The use of negation can be optimized as follows:
\begin{itemize}
\item Constants: $\neg \tautology = \nil$ and $\neg \nil = \tautology$
\item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
\item Double negations: $\neg \neg \psi = \psi$
\end{itemize}

\item 
When forming a product $P(\psi_1, \psi_2, \ldots, \psi_k)$ we can perform the following simplifications:
\begin{itemize}
\item If $\psi_i = \nil$ for any argument $\psi_i$, then the result is $\nil$
\item Eliminate any argument $\psi_i$ for which $\psi_i = \tautology$.
\item If the reduced set of arguments consists of a single formula, then this formula becomes the result.
\item Otherwise, form a product operation over the reduced set of arguments.
\end{itemize}

\item
Similarly, when forming a sum $S(\psi_1, \psi_2, \ldots, \psi_k)$ we can perform the following simplifications:
\begin{itemize}
\item If $\psi_i = \tautology$ for any argument $\psi_i$, then the result is $\tautology$
\item Eliminate any argument $\psi_i$ for which $\psi_i = \nil$.
\item If the reduced set of arguments consists of a single formula, then this formula becomes the result.
\item Otherwise, form a sum operation over the reduced set of arguments.
\end{itemize}
\end{itemize}

@@

We assume that we have a knowledge compiler that can convert a clausal formula $\phi$ into a logically equivalent POF $\psi$.  Furthermore, we assume that the generated POF obeys the following restrictions:
\begin{itemize}
\item The POF is in normal form, i.e., there are no negated product or sum operations.
\item The sum operations are \emph{decision operations}.  That is, each sum operation is of the form $S(\psi_1, \psi_2)$, and it has an associated decision variable $x$,
  such that 1) either $\psi_1 = x$ or $\psi_1$ is a product operation having $x$ as an argument, and 2)
either $\psi_2 = \obar{x}$ or $\psi_2$ is a product operation having $\obar{x}$ as an argument.
\end{itemize}
We let $\kc(\phi)$ denote the output of the knowledge compiler for input formula $\phi$.

@@

Our goal is to implement a projecting knowledge compiler that produces a POF $\pkc(\phi)$ when given input formula $\phi$.
We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that traverses a POF $\psi$ while maintaining its clausal formula $\phi$.  Projected knowledge compilation is then defined
as $\pkc(\phi) = \traverse(\kc(\phi), \bcp(\phi))$.

@@

\item[Literals:]
  $\traverse(\lit, \phi) = \lit$ if $\dependencyset(\lit) \subseteq \dvarset$, and $\traverse(\lit, \phi) = \tautology$ if $\dependencyset(\lit) \in \pvarset$.
\item[Product Operations:]
  Let the operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
  \begin{enumerate}
  \item Let $L = \{\lit_1, \lit_2, \ldots, \lit_k\}$.
  \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.  (Partitioning is only required when $m > 1$.)
  \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
  \end{enumerate}
\item[Sum Operations:]
  Let the operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
  \begin{enumerate}
  \item For $1 \leq i \leq 2$, recursively compute $\psi'_i = \traverse(\psi_i, \phi)$.
  \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
  \item Otherwise, let $\psi' = \pkc(\ureduce{\phi}{x})$
    \begin{enumerate}
    \item If $\psi' = \nil$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$
    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi'), \psi'_2)$.
    \end{enumerate}
  \end{enumerate}
\end{description}

Of these steps, the most interesting ones involve the handling of sum
operations when the decision variable $x$ is to be projected.  The
recursive calls yield subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
These
formulas potentially have overlapping models and so care must be taken
in forming their sum.  Formula $\ureduce{\phi}{x}$ is the clausal
representation of the universal quantification of $\phi$ with respect
to $x$, and so $\psi' = \pkc(\ureduce{\phi}{x})$ will be a POF whose
data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
the event that $\psi' = \nil$, the two formulas are disjoint.  This
will occur, for example, when $x$ is a Tseitin variable encoding some
Boolean formula over the data variables.  In that case, the sum can be
formed directly.  If $\psi' \not = \tautology$, we want to
``subtract'' the overlapping models from one of the arguments before
forming the sum with the other.  We can do so using DeMorgan's Laws.  
That is formula $\neg S(\neg \psi'_1, \psi')$ will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi')$.  Furthermore, the arguments to this sum operation will have disjoint models
since $\dmodelset(\psi') \subseteq \dmodelset(\psi'_1)$, and therefore
$\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
The two arguments to the top-level sum operation are disjoint, because the overlapping models have been subtracted from the first argument.
As can be seen, we could also reverse the role of the two arguments, computing the result
$S(\psi'_1, \neg S(\neg \psi'_2, \psi'))$.

@@

We can improve the efficiency of the traversal by applying the following optimizations:
\begin{itemize}
\item Only data variables in $\psi$.  If $\dependencyset(\psi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \psi$.
\item Only data variables in $\phi$.  If $\dependencyset(\phi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \kc(\phi)$.
\item Only projection variables in $\phi$.  If $\dependencyset(\phi) \subseteq \pvarset$, then $\traverse(\psi, \phi) = \tautology$ if formula $\phi$ is satisfiable and $\nil$ otherwise.
\item Cached result. If $\psi$ is a sum operation and some previous traversal had the same clausal formula $\phi$ as its second argument, then the previously computed result can be reused.\footnote{
Caching for product operations is not, in general valid, since the two recursive calls made when processing a sum operation can be for product operations with the same clausal argument.}
\item Pure literal reduction.
Literal $\lit$ is said to be \emph{pure} in $\phi$ if some clause in $\phi$ contains $\lit$, and no clause in $\phi$ contains  $\obar{\lit}$.
Let $\pure(\phi)$ denote the set of all pure literals in $\phi$, and $\pure_{\pvarset}(\phi) = \{ \lit \in \pure(\phi) \mid  \dependencyset(\lit) \subseteq \pvarset \}$.
If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
$\phi \leftarrow \bcp(\simplify{\phi}{\pure_{\pvarset}(\phi)})$.
\end{itemize}

The first two termination rules have some subtle difference.  For the
initial formulation of $\traverse$, the arguments will always satisfy
$\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, the
pure literal reduction rule can eliminate some literals from $\phi$
that occur in $\psi$, and so the two sets of variables can become
incomparable.  When both rules are applicable, it is more efficient to use the first rule, since it avoids an invocation of $\kc$.

>>>>

\newcommand{\algo}[1]{\textsf{#1}}
\newcommand{\kc}{\algo{KC}}
\newcommand{\pkc}{\algo{PKC}}
\newcommand{\traverse}{\algo{Traverse}}
\newcommand{\cfilter}{\algo{FilterClause}}
\newcommand{\lfilter}{\algo{FilterLiteral}}
\newcommand{\bcp}{\algo{BCP}}
\newcommand{\uprop}{\algo{Uprop}}
\newcommand{\simpl}{\also{Simpl}}

@@

  Many tools for reasoning about Boolean formulas operate on formulas in \emph{conjunctive normal form}, defined as follows.
  A formula is in  \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, ii) the sum operation is applied only to literals, and iii) the product is applied only to sums.  We use symbol $\phi$ to denote CNF formulas.

@@

  A \emph{commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
We write $a - b$ as a shorthand for $a + -b$.

@@

  For Boolean formula $\theta$ and ring weight function $w$, its {\em data ring evaluation} computes
  \begin{eqnarray}
    \drep[\theta, w] & = & \sum_{\alpha \in \dmodelset(\theta)} \;\; \prod_{\lit \in \alpha} w(\lit) \label{eqn:rep}
  \end{eqnarray}
where $\sum$ and $\prod$ represent the ring sum $\radd$ and ring product $\rmul$ operations, respectively.

For a partitioned-operation formula $\psi$ such that $\dependencyset(\psi) \subseteq \dvarset$, its data ring evaluation can be computed based on the recursive structure of the formula:
\begin{itemize}
\item Constants: $\drep[\tautology] = \mulident$ and $\drep[\nil] = \addident$
\item Literals: $\drep[x] = w(x)$ and $\drep[\obar{x}] = \mulident - x$
\item Product operations: 
$\drep[P(\psi_1, \psi_2, \ldots, \psi_k)] = \prod_{1 \leq i \leq k} \drep(\psi_i)$
\item Sum Operations
$\drep[S(\psi_1, \psi_2, \ldots, \psi_k)] = \sum_{1 \leq i \leq k} \drep(\psi_i)$
\item Negations: $\drep[\neg \psi] = \mulident -\drep[\psi]$
\end{itemize}

@@

We can implement a projected knowledge compiler that operates by making multiple invocations of a standard knowledge compiler.

@@

Let $U(\phi)$ denote the set of \emph{unit literals} in $\phi$, i.e., the set of all literals $\lit$ such that clause $\{ \lit \} \in \phi$.  
We define the following operations for transforming clausal formula $\phi$ into a new clausal formula.  In the following $L$ denotes a set of literals.
\begin{description}
\item[$\cfilter(\phi, L)$:] Remove clauses satisfied by the literals in $L$: $\cfilter(\phi, L) = \{ C \in \phi \mid C \cap L = \emptyset\}$.

\item[$\lfilter(\phi, L)$:] Remove the literals in $L$ from all clauses.  
   $\lfilter(\phi, L) = \{ C - L \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.

\item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals in $L$, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
$\simplify{\phi}{L} = \lfilter(\cfilter(\phi, L), \obar{L})$, where $\obar{L} = \{ \obar{\lit} \mid \lit \in L \}$.

@@

\item[$\ureduce{\phi}{x}$:] Express the universal quantification of $\phi$ with respect to variable $x$ in clausal form, by removing all occurrences of $x$ and $\obar{x}$:
$\ureduce{\phi}{x} = \lfilter(\phi, \{x, \obar{x}\})$.

@@

When constructing a POF via a bottom-up process, we can perform simplifications as the formula is constructed:
\begin{itemize}
\item
Optimized negation:
\begin{itemize}
\item Constants: $\neg \tautology = \nil$ and $\neg \nil = \tautology$
\item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
\item Double negations: $\neg \neg \psi = \psi$
\end{itemize}

\item 
Optimizing
product $P(\psi_1, \psi_2, \ldots, \psi_k)$:
\begin{itemize}
\item If $\psi_i = \nil$ for any argument $\psi_i$, then the result is $\nil$
\item Eliminate any argument $\psi_i$ for which $\psi_i = \tautology$.
\item If the reduced set of arguments consists of a single formula, then this formula becomes the result.
\item Otherwise, form a product operation over the reduced set of arguments.
\end{itemize}

\item
Optimizing sum $S(\psi_1, \psi_2, \ldots, \psi_k)$:
\begin{itemize}
\item If $\psi_i = \tautology$ for any argument $\psi_i$, then the result is $\tautology$
\item Eliminate any argument $\psi_i$ for which $\psi_i = \nil$.
\item If the reduced set of arguments consists of a single formula, then this formula becomes the result.
\item Otherwise, form a sum operation over the reduced set of arguments.
\end{itemize}
\end{itemize}

@@

A standard knowledge compiler converts a clausal formula $\phi$ into a logically equivalent POF $\psi$, denoted
$\psi = \kc(\phi)$.
We assume that the generated POF obeys the following restrictions:
\begin{itemize}
\item The POF is in normal form, i.e., there are no negated product or sum operations.
\item The sum operations are \emph{decision operations}.  That is, each sum operation is of the form $S(\psi_1, \psi_2)$, and it has an associated decision variable $x$,
  such that i) either $\psi_1 = x$ or $\psi_1$ is a product operation having $x$ as an argument, and ii)
either $\psi_2 = \obar{x}$ or $\psi_2$ is a product operation having $\obar{x}$ as an argument.
\end{itemize}
These restrictions hold for a knowledge compiler that produces a dec-DNNF representation of the input formula, with this representation converted to a POF\@.

@@

Our goal is to implement a projecting knowledge compiler that produces POF $\psi = \pkc(\phi)$ when given input formula $\phi$.
We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that traverses POF $\psi$ while maintaining its clausal representation $\phi$.  Projected knowledge compilation is then defined
as $\pkc(\phi) = \traverse(\kc(\phi), \bcp(\phi))$.

@@

\item[Literals:]
  $\traverse(\lit, \phi) = \lit$ if $\dependencyset(\lit) \subseteq \dvarset$, and $\traverse(\lit, \phi) = \tautology$ if $\dependencyset(\lit) \in \pvarset$.
\item[Product Operations:]
  Let the operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
  \begin{enumerate}
  \item Let $L = \{\lit_1, \lit_2, \ldots, \lit_k\}$.
  \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.  (Partitioning is only required when $m > 1$.)
  \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
  \end{enumerate}
\item[Sum Operations:]
  Let the operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
  \begin{enumerate}
  \item For $1 \leq i \leq 2$, recursively compute $\psi'_i = \traverse(\psi_i, \phi)$.
  \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
  \item Otherwise, let $\psi_{\forall} = \pkc(\ureduce{\phi}{x})$
    \begin{enumerate}
    \item If $\psi_{\forall} = \nil$, then return the sum (with simplifications) $S(\psi_1, \psi_2)$
    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi_{\forall}), \psi'_2)$.
    \end{enumerate}
  \end{enumerate}
\end{description}

Of these steps, the most interesting ones involve the handling of sum
operations when the decision variable $x$ is to be projected.  The
recursive calls yield subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
These
formulas potentially have overlapping models and so care must be taken
in forming their sum.  Formula $\ureduce{\phi}{x}$ is the clausal
representation of the universal quantification of $\phi$ with respect
to $x$, and so $\psi_{\forall} = \pkc(\ureduce{\phi}{x})$ will be a POF whose
data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
the event that $\psi' = \nil$, the two formulas are disjoint.  This
will occur, for example, when $x$ is a Tseitin variable encoding some
Boolean formula over the data variables.  In that case, the sum can be
formed directly.  If $\psi_{\forall} \not = \nil$, we want to
``subtract'' the overlapping models from one of the arguments before
forming the sum with the other.  We can do so using DeMorgan's Laws.  
That is formula $\neg S(\neg \psi'_1, \psi_{\forall})$ is logically equivalent to $\psi'_1 \land \neg \psi_{\forall}$,
and so it
will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi_{\forall})$.  Furthermore, the arguments to this sum operation will have disjoint models
since $\dmodelset(\psi_{\forall}) \subseteq \dmodelset(\psi'_1)$, and therefore
$\dmodelset(\psi_{\forall}) \cap \dmodelset(\neg \psi'_1) = \emptyset$.
The two arguments to the top-level sum operation are disjoint, because the overlapping models have been subtracted from the first argument.
As can be seen, we could also reverse the role of the two arguments, computing the result
$S(\psi'_1, \neg S(\neg \psi'_2, \psi_{\forall}))$.

@@

We can improve the efficiency of the traversal by applying the following optimizations:
\begin{itemize}
\item Only data variables in $\psi$.  If $\dependencyset(\psi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \psi$.
\item Only data variables in $\phi$.  If $\dependencyset(\phi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \kc(\phi)$.
\item Only projection variables in $\phi$.  If $\dependencyset(\phi) \subseteq \pvarset$, then $\traverse(\psi, \phi) = \tautology$ if formula $\phi$ is satisfiable and $\nil$ otherwise.
\item Cached result. If $\psi$ is a sum operation and some previous traversal had the same clausal formula $\phi$ as its second argument, then the previously computed result can be reused.\footnote{
Caching for product operations is not, in general valid, since the two recursive calls made when processing a sum operation can be for product operations with the same clausal argument.}
\item Pure literal reduction.
Literal $\lit$ is said to be \emph{pure} in $\phi$ if some clause in $\phi$ contains $\lit$, and no clause in $\phi$ contains  $\obar{\lit}$.
Let $\pure(\phi)$ denote the set of all pure literals in $\phi$, and $\pure_{\pvarset}(\phi) = \{ \lit \in \pure(\phi) \mid  \dependencyset(\lit) \subseteq \pvarset \}$.
If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
 $\phi \leftarrow \cfilter(\phi, \pure_{\pvarset})$.  Eliminating clauses can cause other literals to become pure, and so this reduction should be iterated until no pure literals remain.
\end{itemize}

The first two termination rules have some subtle differences.  For the
unoptimized version of $\traverse$, the arguments will always satisfy
$\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, 
pure literal reduction can eliminate some literals from $\phi$
that remain in $\psi$, and so the two sets of variables can become
incomparable.  When both rules are applicable, it is more efficient to use the first, since it avoids an invocation of $\kc$.
